% !TEX program = optex
% !TEX root = ambrojak-diploma-thesis.tex
% The documentation of the usage of CTUstyle -- the template for
% typessetting thesis by plain\TeX at CTU in Prague
% ---------------------------------------------------------------------
% Petr Olsak  Jan. 2013

% You can copy this file to your own file and do some changes.
% Then you can run:  optex your-file

\input ctustyle3  % The template (in version 3, for OpTeX) is included here.

\worktype [M/EN] % Type: B = bachelor, M = master, D = Ph.D., O = other
                 % / the language: CZ = Czech, SK = Slovak, EN = English

\faculty    {F3}  % Type your faculty F1, F2, F3, etc. or MUVS
            % use main language of your document here:
\department {Department of Computer Science}
\title      {Anomaly Detection in Metrans Transport Orders}
% \subtitle   {Software or Research project}
% \subtitleCZ   {Softwarový nebo výzkumný projekt}
            % \subtitle is optional
\author     {Bc. Jakub Ambroz}
\date       {September 2024, May 2025}
\supervisor {Ing. Martin Komárek}  % One or more supervisors
\studyinfo  {Open informatics, specialization Data science}  % Study programme etc.
% \workname   {Software or Research project} % Used only if \worktype [O/*] (Other)
            % optional more information about the document:
\workinfo   {}
            % Title / Subtitle in minor language:
\titleCZ    {Detekce anomálií v transportních objednávkách firmy Metrans}
\subtitleCZ {}
            % If minor language is other than English
            % use \titleCZ, \subtitleCZ or \titleSK, \subtitleSK instead it.
\pagetwo    {}  % The text printed on the page 2 at the bottom.

\abstractEN {
   
}
\abstractCZ {

}
   

\keywordsEN {%
Anomaly detection, 
}
\keywordsCZ {%
Detekce anomálií,
}
\thanks {           % Use main language here
   I would like to thank Stratox, especially Martin Komínek, for his invaluable support in resolving the technical challenges I encountered.
   
   I am also grateful to Metrans for providing real-world data essential to this project, and to their employees Martin Pacourek, Josef Horák and others for their guidance in contextualizing the data and linking it to operational meaning.
}
\declaration {      % Use main language here
    I declare that I have prepared the final thesis independently and have listed all information sources used in accordance with the Methodological Guidelines on the observance of ethical principles in the preparation of university final theses and the Framework Rules for the use of artificial intelligence at CTU for study and pedagogical purposes in Bc and NM studies.
    
    I declare that I have used artificial intelligence tools during the preparation and writing of the final thesis. I have verified the generated content. I confirm that I am aware that I am fully responsible for the content of the final thesis.

   \nl
   (CZECH VERSION)
   
   Prohlašuji, že jsem diplomovou práci vypracoval samostatně a uvedl veškeré použité informační zdroje v souladu s Metodickým pokynem o dodržování
    etických principů při přípravě vysokoškolských závěrečných prací a Rámocovými pravidly používání umělé
    inteligence na ČVUT pro studijní a pedagogické účely v Bc a NM studiu.
    
    Prohlašuji, že jsem v průběhu příprav a psaní závěrečné práce použil nástroje umělé inteligence. Vygenerovaný
    obsah jsem ověřil. Stvrzuji, že jsem si vědom, že za obsah závěrečné práce plně zodpovídám. 

    
    \signature

    In~Prague 23.05.2025
    
    (V~Praze dne 23.05.2025),

    Bc.~Jakub Ambroz
}

%%%%% <--   % The place for your own macros is here.

\draft     % Uncomment this if the version of your document is working only.
%\linespacing=1.7  % uncomment this if you need more spaces between lines
                   % Warning: this works only when \draft is activated!
%\savetoner        % Turns off the lightBlue backround of tables and
                   % verbatims, only for \draft version.
%\blackwhite       % Use this if you need really Black+White thesis.
%\onesideprinting  % Use this if you really don't use duplex printing. 
%\totf             % More compact list of tables and figures.

\input glosdata

\makefront  % Mandatory command. Makes title page, acknowledgment, contents etc.

\input introduction
\input anomalies
\input approaches
\input my_models

% \input uvod    % Files where the source of the document is prepared.
% \input popis   % Full name is: uvod.tex, popis.tex, the suffix can be omitted.
% \input prilohy

\bibchap

\app Glossary
\makeglos

\bye
